\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-08-19 paper.tex
update tilde g notations
minor updates

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty with $\norm{z}_{2,l}=\norm{z^l}_2$, for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g^1_{t_j, \gamma_k}(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g^1_{t, \gamma_k^{(w+1)}}(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
\begin{align}
  \label{eq:update_phi}
  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
\end{align}
Appendix~\ref{sec:?} gives some details about the way we implement the equations and their computation in practice.

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}. 
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$. Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.

@@

\subsection{\texttt{MLMM class}}

@@

\subsection{\texttt{QNMCEM class}}

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto b_i g^4_{s, \gamma_k}(b_i), \\
    \tilde g^6_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}

>>>>

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for the elastic net penalty of any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty of any vector $z$, with $\norm{z}_{2,l}=\norm{z^l}_2$, and for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
\begin{align}
  \label{eq:update_phi}
  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
\end{align}
Appendix~\ref{sec:QNMCEM class} gives some details about the way we implement the equations and their computation in practice.

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}. 
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$. Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.

@@

\subsection{\texttt{MLMM class}}
\label{sec:MLMM class}

@@

\subsection{\texttt{QNMCEM class}}
\label{sec:QNMCEM class}

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}
